Top-down and bottom-up cohesiveness\\in microbial community coalescence

### 2021-06-03 draft/draft_v4.tex
changes to draft v4, adding code for supplementary figure 5 (4?)

\textcolor{red}{To do: figSX, re-write minimal model, check literature, check p-values.}

@@

Microbial communities often invade one another.
This has been observed, for instance, 
in river courses where terrestrial microbes mix with aquatic microorganisms
\cite{Mansour2018,Luo2020,Vass2021},
or in soil communities being invaded as a result of
tillage and outplanting \cite{Rillig2016} or by
aerially dispersed bacteria and funghi \cite{Evans2019}.
Gut microbiomes can invade external communities
through the host's secretions \cite{Dutton2021},
and the skin microbiota is also subject to
invasions when it makes contact with environmental sources of microbes \cite{Vandegrift2019}.

The phenomenon by which entire microbiomes invade one another has been termed
\textit{community coalescence} \cite{Rillig2015}.
Ecologists have long contemplated the idea that interactions between multiple co-invading species
can produce correlated invasional outcomes
\cite{Gilpin1994,Simberloff1999,Grosholz2005,Simberloff2006,Gurevitch2006,Green2011,
Livingston2013,Prior2015,Rillig2015,OLoughlin2017,Castledine2020}.
However, and
in spite of its clear potential importance, the role of coalescence in microbiome assembly is
only beginning to be addressed and little is known about the mechanisms that govern it and its
potential implications.
Early mathematical models of community-community invasions \cite{Gilpin1994,Toquenaga1997}
as well as more recent work
\cite{Tikhonov2016,Tikhonov2017,Vila2019,Lechon2021}
suggest that high-order invasion effects are common
during community coalescence. Communities that have a previous history of coexistence may exhibit an
emergent ``cohesiveness'' which produces correlated invasional outcomes among species from the
same community \cite{Livingston2013,Sierocinski2017}.
The situation where ecological partners in the invading community recruit each other into the final
coalesced community has been called \textit{ecological co-selection} \cite{Rillig2017,Sierocinski2017}.

@@

\bigskip
\begin{center}
\includegraphics[width=11cm,keepaspectratio]{figs/figBox_v3.png}
\end{center}
\bigskip

@@

\begin{equation}
\frac{dR_\alpha}{dt} = 
- \sum_j N_j c_{j\alpha} R_\alpha
+ \sum_j \sum_\beta N_j c_{j\beta} R_\beta
\left[
l_\beta D_{j\alpha\beta} \frac{w_\beta}{w_\alpha}
\right]
\label{eq:crm-r}
\end{equation}
%
These equations can take slightly different forms in certain cases,
e.g. if the primary resource is supplied continuously instead of at the beggining
of each growth cycle \cite{Marsland2019,Marsland2020}.
They represent a good approximation for the community dynamics between consecutive
serial dilutions in our setup.
Here, we assembled \textit{in silico} communities by randomly
sampling a set of species from a pool,
then integrating equations \ref{eq:crm-n} and \ref{eq:crm-r},
diluting the final abundances,
replenishing the primary resource,
and repeating the process until generational equilibrium was achieved
(\methodsref{sim}).
Coalescence simulations were carried out following the same logic,
this time seeding the coalesced communities by mixing the invasive and resident
ones instead of sampling from a species pool.

@@

We then asked whether this trend was also observed in our experimental communities.
To address this question,
we carried out a new round of experiments where we invaded the resident communities
with the invasive dominants alone
(\methodsref{competitions}).
After stabilization
(\methodsref{community-assembly}),
we quantified species abundance through 16S Illumina sequencing
(\methodsref{sequencing}).
Consistent with the behavior of our model,
we observed that bottom-up co-selection is far more common in its
positive than in its negative form
(\figref[d]{fig3}).
Interestingly, bottom-up recruitment appears to be more frequent in the glutamine
environments, where top-down co-selection was weak,
than in the citrate ones,
where top-down co-selection was strong
(\figref{fig2}).
We then repeated our analysis in \figref[c]{fig3},
this time splitting our data according to the observed strength of bottom-up
co-selection instead of the primary carbon source
as we had done in \figref[a]{fig2}.
Our findings were in line with the model prediction:
pairwise competition between dominants is only predictive of coalescence outcomes
if bottom-up co-selection is weak
(\figref[e]{fig3}, $R^2=0.07$, $p>0.05$ when bottom-up co-selection is strong;
$R^2=0.37$, $p<10^{-4}$ when bottom-up co-selection is weak).
Once the bottom-up communities are removed,
the glutamine and citrate communities behave very similarly and both have similar degrees of top-down cohesiveness
(\figref[e]{fig3} right panel,
$R^2=0.27$ for the glutamine environments, in red; $R^2=0.54$ for the citrate environments, in blue;
and $R^2=0.37$ when both environments are considered, black line).
This suggests that the main difference between citrate and glutamine
is that the latter is richer in communities exhibiting bottom-up cohesiveness than the former.

@@

To better understand the underlying mechanisms that govern
the emergence of ecological co-selection in our experiments,
we found it useful to study a minimal model of community coalescence
(\methodsref{sim-min}).
This model is comprised of two communities with two species each
as illustrated in \figref[a]{fig4}.
Within each community, the dominant species
($\mathrm{S}_1$ and $\mathrm{s}_1$ in the resident and invasive communities
respectively)
are able to utilize the single externally supplied resource ($\mathrm{R}_1$).
They secrete a single byproduct
($\mathrm{R}_2$ and $\mathrm{r}_2$ respectively)
off which
the sub-dominants ($\mathrm{S}_2$ and $\mathrm{s}_2$ respectively) can feed.
Finally, these sub-dominants secrete an additional resource
($\mathrm{R}_3$ and $\mathrm{r}_3$ respectively)
that can in turn
be metabolized by the corresponding dominants.
With this structure, community members interact via simple cross-feeding networks that can be
very vertical if the dominant cross-feeds the sub-dominant but the sub-dominant does not cross-feed
the dominant (i.e. if the dominant cannot utilize the sub-dominant's secretions),
or more horizontal (bidirectional) if the cohort also cross-feeds the dominant.
We hypothesize that when two such communities coalesce,
bottom-up ecological co-selection can emerge if the invasive community is
sustained by a bidirectional cross-feeding network.

In the limit case when the cross-feeding networks of both the invasive and resident communities
are strictly vertical (that is, the sub-dominants are passively sustained by the dominants
but do not cross-feed them),
it is straightforward that the outcome of community coalescence will depend on the competitive
ability of the dominants to grow on the single externally supplied resource.
The most competitive dominant will co-select its sub-dominant (from the top-down)
through the secretion of specific metabolic byproducts
(\figref[b]{fig4}).
If the resident community is maintained by a more horizontal cross-feeding network,
it can display further resistance to invasion.
In this scenario, even if the resident dominant is less competitive for the externally supplied resource
than the invasive dominant,
cross-feeding from the resident cohort can favor its success in coalescence.
The stronger the metabolic flux from the resident cohort towards the dominant,
the more prominent this effect can be
(\figref[c]{fig4}).
On the other hand, if the cross-feeding network of the invasive community is horizontal
(i.e. the sub-dominant is cross-fed by and also cross-feeds the dominant),
more complex behaviors can emerge.
The invasive dominant may not be able to invade the resident community by itself
if it is less competitive for the externally supplied resource than the resident dominant
(\figref[d]{fig4}),
or if despite being more competitive, cross-feeding from the resident cohort towards
the resident dominant favors the success of the latter
(\figref[e]{fig4}).
But even then, the invasive community could dominate in coalescence (i.e. when the invasive
sub-dominant is also present).
In order for this to happen, cross-feeding from the invasive cohort towards the invasive dominant
should be strong enough to overcome the competitive disadvantage that said dominant
may have in isolation.

In summary, coalescence outcomes are contingent on the direction of the
cross-feeding networks sustaining the communities in this simple setting.
We ran simulations of all scenarios described above with the minimal model of community coalescence
implemented in the MicroCRM framework
(\methodsref{sim-min}).
In line with our hypothesis, simulations indicate that
bottom-up co-selection of a dominant that is unable to invade by itself
is only possible if said dominant is strongly cross-fed by its cohort
(\figref{fig4}).

@@

\begin{equation}
h = \frac{\Delta N_{\mathrm{dom}}^{\mathrm{R1}}}{\Delta N_{\mathrm{dom}}}
\label{eq:h}
\end{equation}
%
where $\Delta N_{\mathrm{dom}}$ represents the overall increase in dominant biomass
within a single batch incubation for a generationally stable community,
and $\Delta N_{\mathrm{dom}}^{\mathrm{R1}}$
represents the increase in said biomass resulting from the metabolism of the primary resource
($\mathrm{R}_1$) only.
If the dominant was just utilizing the primary resource, the cross-feeding network would be very
hierarchical ($h \sim 1$),
whereas if it was growing mostly on the secretions of other taxa,
it would be more distributed ($h << 1$).
We quantified the hierarchies of the resident and invasive communities in our MicroCRM,
finding that $h$ follows a bimodal distribution (\figref[a]{fig5}).
We therefore divided our simulations into four groups according to
whether the cross-feeding networks of both resident and invasive communities were
vertical (high $h$) or horizontal (low $h$) as shown in
\figref[b]{fig5}.
For each group, we evaluated the frequency of instances of bottom-up co-selection,
i.e. the fraction of cases where a dominant that could not invade in isolation
was successful when accompanied by its cohort
(green area of \figref[b]{fig3}).
We found that bottom-up ecological co-selection
is significantly more frequent when the invasive community is
non-hierarchical (\figref[c]{fig5}),
in line with what the minimal model anticipated
(\figref[d-e]{fig4}).

@@

>>>>

@@

Microbial communities often invade one another.
This has been observed, for instance, 
in river courses where terrestrial microbes mix with aquatic microorganisms
\cite{Mansour2018,Luo2020,Vass2021},
or in soil communities being invaded as a result of
tillage and outplanting
\cite{Rillig2016,Ramoneda2020,Rochefort2020}
or by aerially dispersed bacteria and funghi \cite{Evans2019}.
Gut microbiomes can invade external communities
through the host's secretions \cite{Dutton2021},
and the skin microbiota is also subject to
invasions when it makes contact with environmental sources of microbes \cite{Vandegrift2019}.

The phenomenon by which entire microbiomes invade one another has been termed
\textit{community coalescence} \cite{Rillig2015}.
Ecologists have long contemplated the idea that interactions between multiple co-invading species
can produce correlated invasional outcomes
\cite{Gilpin1994,Simberloff1999,Grosholz2005,Simberloff2006,Gurevitch2006,Green2011,
Livingston2013,Prior2015,Rillig2015,OLoughlin2017,Castledine2020}.
However, and
in spite of its clear potential importance, the role of coalescence in microbiome assembly is
only beginning to be addressed and little is known about the mechanisms that govern it and its
potential implications or applications
\cite{Rillig2016b}.
Early mathematical models of community-community invasions \cite{Gilpin1994,Toquenaga1997}
as well as more recent work
\cite{Tikhonov2016,Tikhonov2017,Vila2019,Lechon2021}
suggest that high-order invasion effects are common
during community coalescence. Communities that have a previous history of coexistence may exhibit an
emergent ``cohesiveness'' which produces correlated invasional outcomes among species from the
same community \cite{Livingston2013,Sierocinski2017}.
The situation where ecological partners in the invading community recruit each other into the final
coalesced community has been called \textit{ecological co-selection} \cite{Rillig2017,Sierocinski2017}.

@@

\bigskip
\begin{center}
\includegraphics[width=11cm,keepaspectratio]{figs/figBox_v4.png}
\end{center}
\bigskip

@@

\begin{equation}
\frac{dR_\alpha}{dt} = 
- \sum_j N_j c_{j\alpha} R_\alpha
+ \sum_j \sum_\beta N_j c_{j\beta} R_\beta
\left[
l_\beta D_{j\alpha\beta} \frac{w_\beta}{w_\alpha}
\right]
\label{eq:crm-r}
\end{equation}
%
These equations can take slightly different forms in certain cases,
e.g. if the primary resource is supplied continuously instead of at the beginning
of each growth cycle \cite{Marsland2019,Marsland2020}.
They represent a good approximation for the community dynamics between consecutive
serial dilutions in our setup.
Here, we assembled \textit{in silico} communities by randomly
sampling a set of species from a pool,
then integrating equations \ref{eq:crm-n} and \ref{eq:crm-r},
diluting the final abundances,
replenishing the primary resource,
and repeating the process until generational equilibrium was achieved
(\methodsref{sim}).
Coalescence simulations were carried out following the same logic,
this time seeding the coalesced communities by mixing the invasive and resident
ones instead of sampling from a species pool.

@@

We then asked whether this trend was also observed in our experimental communities.
To address this question,
we carried out a new round of experiments where we invaded the resident communities
with the invasive dominants alone
(\methodsref{competitions}).
After stabilization
(\methodsref{community-assembly}),
we quantified species abundance through 16S Illumina sequencing
(\methodsref{sequencing}).
Consistent with the behavior of our model,
we observed that bottom-up co-selection is far more common in its
positive than in its negative form
(\figref[d]{fig3}).
Interestingly, bottom-up recruitment appears to be more frequent in the glutamine
environments, where top-down co-selection was weak,
than in the citrate ones,
where top-down co-selection was strong
(\figref{fig2}).
We then repeated our analysis in \figref[c]{fig3},
this time splitting our data according to the observed strength of bottom-up
co-selection instead of the primary carbon source
as we had done in \figref[a]{fig2}.
Our findings were in line with the model prediction:
pairwise competition between dominants is only predictive of coalescence outcomes
if bottom-up co-selection is weak
(\figref[e]{fig3}, $R^2=0.07$, $p>0.05$ when bottom-up co-selection is strong;
$R^2=0.37$, $p<10^{-4}$ when bottom-up co-selection is weak).
Once the bottom-up communities are removed,
both the glutamine and citrate communities
display similar degrees of top-down cohesiveness
(\figref[e]{fig3}, right panel).
This suggests that the main difference between citrate and glutamine
is that the latter is richer in communities exhibiting bottom-up cohesiveness than the former.

@@

In view of the success of our model in reproducing the
experimentally observed trends in ecological co-selection,
we set out to better understand the mechanisms for its emergence.
In our experimental conditions and in the MicroCRM,
comunities are sustained by dense cross-feeding facilitation networks.
These networks can have a very vertical direction if
a single species (the dominant) cross-feeds
the rarer members of the community
but these do not cross feed the dominant in return.
Alternatively, if the dominant is strongly cross-fed by its cohort
the network structure would be more horizontal.
In the latter scenario, positive bottom-up co-selection of a dominant
can take place if
cross-feeding from its cohort allow it
to persist in the final community after coalescence
--even if it cannot invade succesfully in isolation.

We found it useful to study a minimal model of community coalescence
to test these ideas
(\methodsref{sim-min}).
This model is comprised of two communities
(resident and invasive)
with only two species each
as illustrated in \figref[a]{fig4}.
Within each community, the dominant species
($\mathrm{S}_1$ and $\mathrm{s}_1$ respectively)
are able to utilize the single externally supplied resource ($\mathrm{R}_1$).
They secrete a single byproduct
($\mathrm{R}_2$ and $\mathrm{r}_2$ respectively)
off which
the sub-dominants ($\mathrm{S}_2$ and $\mathrm{s}_2$ respectively) can feed.
Finally, these sub-dominants secrete an additional resource
($\mathrm{R}_3$ and $\mathrm{r}_3$ respectively)
that can in turn
be metabolized by the corresponding dominants.
The dominants' ability to utilize their sub-dominants' metabolic
byproducts determines whether the structure of the cross-feeding
networks of these minimal communities is vertical
(if the dominants cannot utilize the cohort secretions
and thus are not cross-fed by them)
or horizontal
(in the opposite scenario).
The model parameters controlling how effectively the dominants
can metabolize said byproducts modulate the direction of the
cross-feeding networks
(\figref[a]{fig4}).

In the limit case when the cross-feeding networks of both the invasive and resident communities
are strictly vertical (that is, the sub-dominants are passively sustained by the dominants
but do not cross-feed them),
it is straightforward that the outcome of community coalescence will depend on the competitive
ability of the dominants to grow on the single externally supplied resource.
The most competitive dominant will co-select its sub-dominant (from the top-down)
through the secretion of specific metabolic byproducts
(\figref[b]{fig4}).
If the resident community is maintained by a more horizontal cross-feeding network,
it can display further resistance to invasion.
In this scenario, even if the resident dominant is less competitive for the externally supplied resource
than the invasive dominant,
cross-feeding from the resident cohort can allow it to persist in coalescence.
The stronger the metabolic flux from the resident cohort towards the dominant,
the more prominent this effect can be
(\figref[c]{fig4}).
On the other hand, if the cross-feeding network of the invasive community is horizontal
(i.e. the sub-dominant is cross-fed by and also cross-feeds the dominant),
more complex behaviors can emerge.
The invasive dominant may not be able to invade the resident community by itself
if it is less competitive for the externally supplied resource than the resident dominant
(\figref[d]{fig4}),
or if despite being more competitive, cross-feeding from the resident cohort towards
the resident dominant favors the success of the latter
(\figref[e]{fig4}).
But even then, the invasive community could dominate in coalescence (i.e. when the invasive
sub-dominant is also present).
In order for this to happen, cross-feeding from the invasive cohort towards the invasive dominant
should be strong enough to overcome the competitive disadvantage that said dominant
may have in isolation.

In summary, coalescence outcomes are contingent on the direction of the
cross-feeding networks sustaining the communities in this simple setting.
We ran simulations of all scenarios described above with our minimal model of community coalescence
implemented in the MicroCRM framework
(\methodsref{sim-min}).
In line with our initial proposition, simulations indicate that
bottom-up co-selection of a dominant that is unable to invade by itself
is possible if said dominant is strongly cross-fed by its cohort
(\figref{fig4}).

@@

\begin{equation}
h = \frac{\Delta N_{\mathrm{dom}}^{\mathrm{R1}}}{\Delta N_{\mathrm{dom}}}
\label{eq:h}
\end{equation}
%
where $\Delta N_{\mathrm{dom}}$ represents the overall increase in dominant biomass
within a single batch incubation for a generationally stable community,
and $\Delta N_{\mathrm{dom}}^{\mathrm{R1}}$
represents the increase in said biomass resulting from the metabolism of the primary resource
($\mathrm{R}_1$) only.
If the dominant was just utilizing the primary resource, the cross-feeding network would be very
vertical ($h \sim 1$),
whereas if it was growing mostly on the secretions of other taxa,
it would be more horizontal ($h << 1$).
We quantified the hierarchies of the resident and invasive communities in our MicroCRM,
finding that $h$ follows a bimodal distribution (\figref[a]{fig5}).
We therefore divided our simulations into four groups according to
whether the cross-feeding networks of both resident and invasive communities were
vertical (high $h$) or horizontal (low $h$) as shown in
\figref[b]{fig5}.
For each group, we evaluated the frequency of instances of bottom-up co-selection,
i.e. the fraction of cases where a dominant that could not invade in isolation
was successful when accompanied by its cohort
(green area of \figref[b]{fig3}).
We found that bottom-up ecological co-selection
is significantly more frequent when the invasive community is
non-hierarchical (\figref[c]{fig5}),
in line with what the minimal model anticipated
(\figref[d-e]{fig4}).

@@

\begin{figure}[!h]
\centering
\internallinenumbers
\includegraphics[scale=0.9,keepaspectratio]{figs/figS4.pdf}
\caption{\textbf{Top-down co-selection is robust in the MicroCRM.}
We ran simulations of community coalescence following the same
procedure described in the main text, but randomly sampling the
following prameters of the MicroCRM
uniformly within the indicated ranges:
number of species per community sampled at initialization ($S$) between 10 and 90;
total number of species families between 1 and 5;
total number of species per family ($S_A$) between 400 and 1200;
total number of generalist species ($S_{gen}$) between 100 and 380;
specialist species preference strength ($q$) between 0.5 and 1;
number of resource classes between 1 and 5;
number of resources per class ($M_A$) between 3 and 17;
leakage fraction ($l_{\alpha}$) between 0.45 and 0.95;
maintenance cost ($m_i$) between 0 and 0.2;
standard deviation of the sum of consumption rates ($\sigma_c$) between 2 and 4;
sparsity of the metabolic matrix between 0.05 and 0.95;
fraction of secretion flux to resources of the same type ($f_s$) or to waste resources ($f_w$) between 0.05 and 0.45.
The metabolic matrix was randomly chosen to be such that $D_{i\beta\alpha}=D_{j\beta\alpha}$ or such that
$D_{i\beta\alpha} \neq D_{j\beta\alpha}$ for all species ($i$, $j$) and resources ($\alpha$, $\beta$).
Both options were given equal probabilities.
For each randomly sampled set of parameters, we ran 100 simulations
and quantified the ability of the pairwise competition
of dominants to predict coalescence outcomes (see \figref[a]{fig2}), i.e. the strength of top-down co-selection
in each regime.
The histogram shows that top-down co-selection is robust throughout the parameter space in the MicroCRM.}
\label{figSX}
\end{figure}
